The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
change definition of cluster (now: induced by node, not by edge)

A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

\todo{define splits and delete definition from split thm chapter!}
\todo{change definition of clusters: induced by vertices, not edges!}

Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
The vertex set of the \emph{$\rnni$-graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space} or just $\rnni$.

@@

\begin{algorithm}[H]
\caption{FIND\_PATH($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with $rank(v) = i < rank(w)$
		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ given by internal edges of $R$, represented by $i = 1, \dots, n-2$ in the algorithm, requires $n-1-i$ $\rnni$ moves for moving $mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

%This part is mainly for myself, not sure whether it is important to have it in the paper
\todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
The counterexample presented in~\cite{Li1996} works in $\nni$ because of the following:
When proving Theorem~\ref{thm:split_nni},~\cite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The clusters given by the two edges incident to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ where $T|_{\{1, \ldots, n\}}$ and $R|_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

@@

Instead of using the notion of clusters as we do within this paper, the Split Theorem considers partitions of the taxon set $X$, so-called splits, received by deleting an edge of a tree and taking the taxon sets of the two resulting trees as partition of $X$.
However, we found a simple counterexample to this conjecture:

>>>>

A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \to \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
Each internal node $v$ gives a \emph{cluster} $C$ that is the set of taxa descending from this node. \todo{OK?}
We then say that $v$ induces the cluster $C$.
Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

\todo{define splits and delete definition from split thm chapter!}

Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space} or just $\rnni$.

@@

\begin{algorithm}[H]
\caption{FIND\_PATH($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster node $v$ in $R$ with $rank(v) = i$
		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $mrca(C)$ down in $\hat{T}$.
	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

%This part is mainly for myself, not sure whether it is important to have it in the paper
\todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
The counterexample presented in~\cite{Li1996} works in $\nni$ because of the following:
When proving Theorem~\ref{thm:split_nni},~\cite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ where $T|_{\{1, \ldots, n\}}$ and $R|_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

@@

Instead of using the notion of clusters as we do within this paper, the Split Theorem considers splits of the taxon set $X$.
However, we found a simple counterexample to this conjecture:
